---
title: "00 EDA Overview: core warehouse profiling"
subtitle: "Independent read only profile of every dim and fact before targeted analysis"
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
execute:
  warning: false
  message: false
---

## Purpose

This notebook profiles every table in the `core` schema before any targeted
analysis is written, so every later notebook starts from known, checked data.
It is independent confirmation from the R side. It does not replace the five
PL/pgSQL quality loops. It only reads. It creates no tables, no views, no
procedures.

Render it on demand from the repository root:

```bash
make eda
```

The render needs a populated warehouse and a `.env` file with the same
`POSTGRES_*` values the Python pipeline uses. Rendered HTML stays out of git
by design. Rerun this notebook whenever the warehouse grain, keys, or load
logic changes.

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages({
  library(DBI)
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(skimr)
  library(DataExplorer)
  library(naniar)
  library(GGally)
  library(lubridate)
  library(scales)
  library(knitr)
})

helper_path <- file.path("..", "r_analysis", "scripts", "db_connect.R")
if (!file.exists(helper_path)) {
  helper_path <- file.path("r_analysis", "scripts", "db_connect.R")
}
source(helper_path)

theme_path <- file.path(dirname(helper_path), "plot_theme.R")
source(theme_path)

con <- connect_db()

DIM_TABLES <- c(
  "dim_campaign"
  , "dim_customers"
  , "dim_geo"
  , "dim_orders_flag"
  , "dim_products"
)
FACT_TABLES <- c(
  "fact_campaign_spend"
  , "fact_inventory"
  , "fact_less_fact"
  , "fact_order_process"
  , "fact_orders"
)
SAMPLE_N <- 2000L
```

```{r}
#| label: helpers
#| include: false
table_row_count <- function(con, table) {
  sql <- paste0(
    "SELECT COUNT(*) AS row_count FROM core.", table
  )
  DBI::dbGetQuery(con, sql)$row_count[[1L]]
}

table_columns <- function(con, table) {
  sql <- paste0(
    "SELECT "
    , " c.column_name"
    , " , c.data_type"
    , " , c.is_nullable"
    , " FROM information_schema.columns AS c"
    , " WHERE c.table_schema = 'core'"
    , " AND c.table_name = '", table, "'"
    , " ORDER BY c.ordinal_position"
  )
  DBI::dbGetQuery(con, sql)
}

quoted_ident <- function(x) {
  paste0('"', gsub('"', '""', x, fixed = TRUE), '"')
}

sample_table <- function(con, table, n = SAMPLE_N) {
  cols <- table_columns(con, table)$column_name
  col_list <- paste(vapply(cols, quoted_ident, character(1L)), collapse = " , ")
  sql <- paste0(
    "SELECT "
    , col_list
    , " FROM core.", table
    , " ORDER BY random()"
    , " LIMIT ", as.integer(n)
  )
  DBI::dbGetQuery(con, sql)
}

null_counts <- function(con, table) {
  cols <- table_columns(con, table)$column_name
  parts <- vapply(
    cols
    , function(col) {
      paste0(
        "COUNT(*) FILTER (WHERE "
        , quoted_ident(col), " IS NULL) AS "
        , quoted_ident(paste0(col, "__nulls"))
      )
    }
    , character(1L)
  )
  sql <- paste0(
    "SELECT COUNT(*) AS rows_checked , "
    , paste(parts, collapse = " , ")
    , " FROM core.", table
  )
  DBI::dbGetQuery(con, sql)
}
```

## Headline: shape of the warehouse

Row counts for all ten modelled tables plus the pipeline log. Detail for
each table follows below. If every count below looks plausible and the
verdict section at the end reports zero failures, the warehouse is ready
for targeted analysis.

```{r}
#| label: headline-counts
headline_tables <- c(DIM_TABLES, FACT_TABLES, "pipeline_run_log")
headline <- tibble::tibble(
  table_name = headline_tables
  , row_count = vapply(
    headline_tables
    , function(t) {
      if (t == "pipeline_run_log") {
        exists <- DBI::dbGetQuery(
          con
          , paste0(
            "SELECT COUNT(*) AS n FROM information_schema.tables"
            , " WHERE table_schema = 'core' AND table_name = 'pipeline_run_log'"
          )
        )$n[[1L]] > 0L
        if (!exists) {
          return(NA_real_)
        }
      }
      as.numeric(table_row_count(con, t))
    }
    , numeric(1L)
  )
)
knitr::kable(headline, caption = "Row counts across core")
```

## Dimensions

### Column inventory

One row per column for every dimension: name, type, and nullability. This
is the contract later notebooks rely on. Any drift here means the catalog
in `docs/data_catlog.md` needs an update in the same change.

```{r}
#| label: dim-columns
dim_inventory <- dplyr::bind_rows(lapply(
  DIM_TABLES
  , function(t) {
    cols <- table_columns(con, t)
    cols$table_name <- t
    cols[, c("table_name", "column_name", "data_type", "is_nullable")]
  }
))
knitr::kable(dim_inventory, caption = "Dimension column inventory")
```

### Skim profiles

`skimr::skim()` over a random sample of each dimension. One glance at type,
missing share, and basic stats per column. Full table scans would be slow
on large dimensions, so the sample keeps renders fast while the aggregate
SQL checks further below cover every row.

```{r}
#| label: dim-skim
#| results: asis
for (t in DIM_TABLES) {
  cat("\n\n#### ", t, "\n\n", sep = "")
  s <- sample_table(con, t)
  cat(
    knitr::kable(
      tibble::as_tibble(skimr::skim(s))
      , digits = 2L
      , caption = paste("skim profile:", t)
    )
    , sep = "\n"
  )
  cat("\n")
}
```

### Primary key and unique key checks

Catalog driven duplicate check over every primary and unique constraint in
`core`, the same definition the fourth quality loop enforces. For
`dim_customers` this covers both the surrogate primary key and the SCD2
business key on customer plus version start. A nonzero count here is a
failed check.

```{r}
#| label: dim-key-checks
key_constraints <- DBI::dbGetQuery(
  con
  , paste0(
    "SELECT "
    , " n.nspname AS schema_name"
    , " , c.relname AS table_name"
    , " , con.conname AS constraint_name"
    , " , STRING_AGG(a.attname, ',' ORDER BY k.ordinality) AS columns_csv"
    , " FROM pg_catalog.pg_constraint AS con"
    , " JOIN pg_catalog.pg_class AS c ON c.oid = con.conrelid"
    , " JOIN pg_catalog.pg_namespace AS n ON n.oid = c.relnamespace"
    , " JOIN LATERAL UNNEST(con.conkey)"
    , " WITH ORDINALITY AS k(attnum, ordinality) ON TRUE"
    , " JOIN pg_catalog.pg_attribute AS a"
    , " ON a.attrelid = c.oid AND a.attnum = k.attnum"
    , " WHERE n.nspname = 'core'"
    , " AND c.relname IN ("
    , paste(paste0("'", DIM_TABLES, "'"), collapse = " , ")
    , ")"
    , " AND con.contype IN ('p', 'u')"
    , " GROUP BY n.nspname, c.relname, con.conname"
    , " ORDER BY c.relname, con.conname"
  )
)

dup_results <- dplyr::bind_rows(lapply(
  seq_len(nrow(key_constraints))
  , function(i) {
    row <- key_constraints[i, ]
    cols <- strsplit(row$columns_csv, ",", fixed = TRUE)[[1L]]
    quoted <- vapply(cols, quoted_ident, character(1L))
    group_list <- paste(quoted, collapse = " , ")
    not_null <- paste(paste0(quoted, " IS NOT NULL"), collapse = " AND ")
    sql <- paste0(
      "SELECT COALESCE(SUM(dup_rows), 0) AS duplicate_rows FROM ("
      , " SELECT COUNT(*) - 1 AS dup_rows"
      , " FROM core.", row$table_name
      , " WHERE ", not_null
      , " GROUP BY ", group_list
      , " HAVING COUNT(*) > 1"
      , ") AS g"
    )
    n <- DBI::dbGetQuery(con, sql)$duplicate_rows[[1L]]
    data.frame(
      table_name = row$table_name
      , constraint_name = row$constraint_name
      , duplicate_rows = as.numeric(n)
      , stringsAsFactors = FALSE
    )
  }
))
knitr::kable(dup_results, caption = "Duplicate key rows per dimension constraint")
```

### Categorical columns

Distinct counts for every text column, plus a top values bar chart wherever
cardinality is low enough to plot. High cardinality columns such as names
and emails are counted but not charted, since a chart with hundreds of bars
carries no insight.

```{r}
#| label: dim-categoricals
#| results: asis
for (t in DIM_TABLES) {
  cat("\n\n#### ", t, "\n\n", sep = "")
  cols <- table_columns(con, t)
  text_cols <- cols$column_name[cols$data_type %in% c(
    "character varying", "character", "text"
  )]
  summary_rows <- lapply(
    text_cols
    , function(col) {
      sql <- paste0(
        "SELECT "
        , " COUNT(DISTINCT ", quoted_ident(col), ") AS distinct_count"
        , " , COUNT(*) AS row_count"
        , " FROM core.", t
      )
      res <- DBI::dbGetQuery(con, sql)
      data.frame(
        column_name = col
        , distinct_count = as.numeric(res$distinct_count[[1L]])
        , row_count = as.numeric(res$row_count[[1L]])
        , stringsAsFactors = FALSE
      )
    }
  )
  summary_tbl <- dplyr::bind_rows(summary_rows)
  cat(
    knitr::kable(summary_tbl, caption = paste("Categorical summary:", t))
    , sep = "\n"
  )
  cat("\n")

  for (col in text_cols) {
    dc <- summary_tbl$distinct_count[summary_tbl$column_name == col]
    if (length(dc) == 1L && !is.na(dc) && dc <= 15L && dc > 0L) {
      sql <- paste0(
        "SELECT "
        , quoted_ident(col), " AS value"
        , " , COUNT(*) AS n"
        , " FROM core.", t
        , " GROUP BY ", quoted_ident(col)
        , " ORDER BY COUNT(*) DESC"
        , " LIMIT 10"
      )
      top_vals <- fix_int64(DBI::dbGetQuery(con, sql))
      top_vals$value[is.na(top_vals$value)] <- "NULL"
      p <- ggplot2::ggplot(
        top_vals
        , ggplot2::aes(x = stats::reorder(value, n), y = n)
      ) +
        ggplot2::geom_col(fill = "#3498DB") +
        ggplot2::coord_flip() +
        ggplot2::labs(
          title = paste0(t, ": top values of ", col)
          , x = col
          , y = "Row count"
        )
      print(apply_plot_theme(p))
    }
  }
}
```

### dim_customers SCD2 sanity

The customer dimension keeps history: one open row per customer plus closed
rows for prior values. This section counts open against closed rows, shows
the version count distribution, and summarises version lifespan in days.
Open rows use the current timestamp as their provisional end, so their
spans only grow over time.

```{r}
#| label: dim-customers-scd2
scd2_status <- DBI::dbGetQuery(
  con
  , paste0(
    "SELECT "
    , " COUNT(*) FILTER (WHERE d.is_current) AS open_rows"
    , " , COUNT(*) FILTER (WHERE NOT d.is_current) AS closed_rows"
    , " , COUNT(DISTINCT d.customer_id) AS distinct_customers"
    , " FROM core.dim_customers AS d"
  )
)
knitr::kable(scd2_status, caption = "SCD2 open against closed rows")

scd2_versions <- fix_int64(DBI::dbGetQuery(
  con
  , paste0(
    "SELECT "
    , " d.customer_id"
    , " , COUNT(*) AS version_count"
    , " , MIN(d.valid_from) AS first_valid_from"
    , " , MAX(COALESCE(d.valid_to, now())) AS last_valid_end"
    , " FROM core.dim_customers AS d"
    , " GROUP BY d.customer_id"
    , " ORDER BY COUNT(*) DESC"
  )
))
knitr::kable(
  utils::head(scd2_versions, 15L)
  , caption = "Customers with the most versions, top fifteen"
)

scd2_spans <- DBI::dbGetQuery(
  con
  , paste0(
    "SELECT "
    , " EXTRACT(EPOCH FROM (COALESCE(d.valid_to, now()) - d.valid_from)) / 86400.0"
    , " AS lifespan_days"
    , " , d.is_current"
    , " FROM core.dim_customers AS d"
  )
)
knitr::kable(
  data.frame(
    group = c("all versions", "open rows", "closed rows")
    , min_days = c(
      min(scd2_spans$lifespan_days, na.rm = TRUE)
      , min(scd2_spans$lifespan_days[scd2_spans$is_current], na.rm = TRUE)
      , min(scd2_spans$lifespan_days[!scd2_spans$is_current], na.rm = TRUE)
    )
    , median_days = c(
      stats::median(scd2_spans$lifespan_days, na.rm = TRUE)
      , stats::median(
        scd2_spans$lifespan_days[scd2_spans$is_current], na.rm = TRUE
      )
      , stats::median(
        scd2_spans$lifespan_days[!scd2_spans$is_current], na.rm = TRUE
      )
    )
    , max_days = c(
      max(scd2_spans$lifespan_days, na.rm = TRUE)
      , max(scd2_spans$lifespan_days[scd2_spans$is_current], na.rm = TRUE)
      , max(scd2_spans$lifespan_days[!scd2_spans$is_current], na.rm = TRUE)
    )
  )
  , digits = 1L
  , caption = "Version lifespan in days"
)

p_versions <- ggplot2::ggplot(
  scd2_versions
  , ggplot2::aes(x = version_count)
) +
  ggplot2::geom_histogram(binwidth = 1L, fill = "#3498DB", color = "#2C3E50") +
  ggplot2::labs(
    title = "Customer version count distribution"
    , x = "Versions per customer"
    , y = "Customer count"
  )
apply_plot_theme(p_versions)

p_spans <- ggplot2::ggplot(scd2_spans, ggplot2::aes(x = lifespan_days)) +
  ggplot2::geom_histogram(bins = 40L, fill = "#2ECC71", color = "#2C3E50") +
  ggplot2::facet_wrap(~is_current, labeller = ggplot2::label_both) +
  ggplot2::labs(
    title = "Version lifespan distribution, open against closed"
    , x = "Lifespan in days"
    , y = "Version count"
  )
apply_plot_theme(p_spans)
```

## Facts

### Row counts and grain checks

Each fact declares one natural key in the catalog. The grain check counts
rows that share a key value beyond the first occurrence. Any nonzero count
means the load produced more rows per key than the grain allows, which
would silently multiply measures in downstream joins.

```{r}
#| label: fact-grain
grain_keys <- list(
  fact_orders = c("order_id", "line_id")
  , fact_order_process = c("order_id")
  , fact_inventory = c("product_name", "period_month")
  , fact_campaign_spend = c("campaign_name", "spend_date")
  , fact_less_fact = c("campaign_name", "promoted_sku")
)

grain_results <- dplyr::bind_rows(lapply(
  FACT_TABLES
  , function(t) {
    key_cols <- grain_keys[[t]]
    quoted <- vapply(key_cols, quoted_ident, character(1L))
    group_list <- paste(quoted, collapse = " , ")
    not_null <- paste(paste0(quoted, " IS NOT NULL"), collapse = " AND ")
    sql <- paste0(
      "SELECT "
      , " COUNT(*) AS row_count"
      , " , COALESCE(SUM(dup_rows), 0) AS extra_rows"
      , " FROM core.", t
      , " LEFT JOIN ("
      , " SELECT COUNT(*) - 1 AS dup_rows"
      , " FROM core.", t
      , " WHERE ", not_null
      , " GROUP BY ", group_list
      , " HAVING COUNT(*) > 1"
      , ") AS g ON TRUE"
    )
    res <- DBI::dbGetQuery(con, sql)
    data.frame(
      table_name = t
      , grain = paste(key_cols, collapse = " + ")
      , row_count = as.numeric(res$row_count[[1L]])
      , extra_rows = as.numeric(res$extra_rows[[1L]])
      , stringsAsFactors = FALSE
    )
  }
))
knitr::kable(grain_results, caption = "Fact grain check, extra rows per key")
```

### Missing values

Aggregate NULL counts per column across every row, plus a visual missing
map over a sample for quick scanning. Identifier and key columns are
included here because an unresolved dimension key shows up as NULL on the
fact, and the orphan check later treats that as a failure.

```{r}
#| label: fact-missing
#| results: asis
for (t in FACT_TABLES) {
  cat("\n\n#### ", t, "\n\n", sep = "")
  nc <- null_counts(con, t)
  rows_checked <- nc$rows_checked[[1L]]
  long <- data.frame(
    column_name = sub("__nulls$", "", names(nc)[-1L])
    , null_count = as.numeric(unlist(nc[1L, -1L], use.names = FALSE))
    , stringsAsFactors = FALSE
  )
  long$null_share <- long$null_count / rows_checked
  long <- long[order(-long$null_count), ]
  cat(
    knitr::kable(
      long
      , digits = 4L
      , caption = paste("NULL counts:", t)
    )
    , sep = "\n"
  )
  cat("\n")

  s <- sample_table(con, t, n = 500L)
  print(naniar::vis_miss(s) + ggplot2::labs(title = paste("Missing map:", t)))
  print(DataExplorer::plot_missing(
    s
    , title = paste("Missing profile:", t)
    , ggtheme = get_plot_theme()
  ))
}
```

### Numeric measures

Min, max, mean, and median for every numeric measure, computed over all
rows in SQL. Distribution histograms come from a sample in R. Outliers use
the interquartile rule on the sample: values outside one and a half times
the interquartile range past either quartile. Counts and boundary values
are reported. No business rows are printed.

```{r}
#| label: fact-numerics
#| results: asis
MEASURE_COLS <- list(
  fact_orders = c(
    "quantity", "unit_price", "unit_cost", "discount_pct", "line_total"
  )
  , fact_order_process = c(
    "amount"
    , "days_order_to_ship"
    , "days_ship_to_delivery"
    , "days_order_to_invoice"
    , "days_invoice_to_pay"
  )
  , fact_inventory = c("quantity")
  , fact_campaign_spend = c("impressions", "clicks", "spend")
  , fact_less_fact = character(0L)
)

flag_iqr <- function(x) {
  x <- x[!is.na(x)]
  if (length(x) < 5L) {
    return(list(n = 0L, lower = NA_real_, upper = NA_real_, flagged = 0L))
  }
  q <- stats::quantile(x, probs = c(0.25, 0.75))
  iqr <- q[[2L]] - q[[1L]]
  lower <- q[[1L]] - 1.5 * iqr
  upper <- q[[2L]] + 1.5 * iqr
  list(
    n = length(x)
    , lower = unname(lower)
    , upper = unname(upper)
    , flagged = sum(x < lower | x > upper)
  )
}

for (t in FACT_TABLES) {
  measures <- MEASURE_COLS[[t]]
  cat("\n\n#### ", t, "\n\n", sep = "")
  if (length(measures) == 0L) {
    cat("\nThis fact carries no numeric measures by design.\n\n")
    next
  }
  existing <- intersect(measures, table_columns(con, t)$column_name)
  stat_parts <- vapply(
    existing
    , function(col) {
      q <- quoted_ident(col)
      paste0(
        "MIN(", q, ") AS ", quoted_ident(paste0(col, "__min"))
        , " , MAX(", q, ") AS ", quoted_ident(paste0(col, "__max"))
        , " , AVG(", q, ") AS ", quoted_ident(paste0(col, "__mean"))
        , " , PERCENTILE_CONT(0.5) WITHIN GROUP (ORDER BY ", q, ")"
        , " AS ", quoted_ident(paste0(col, "__median"))
      )
    }
    , character(1L)
  )
  sql <- paste0("SELECT ", paste(stat_parts, collapse = " , "), " FROM core.", t)
  stats_wide <- DBI::dbGetQuery(con, sql)
  stats_long <- data.frame(
    measure = existing
    , min = vapply(
      existing
      , function(col) as.numeric(stats_wide[[paste0(col, "__min")]])
      , numeric(1L)
    )
    , max = vapply(
      existing
      , function(col) as.numeric(stats_wide[[paste0(col, "__max")]])
      , numeric(1L)
    )
    , mean = vapply(
      existing
      , function(col) as.numeric(stats_wide[[paste0(col, "__mean")]])
      , numeric(1L)
    )
    , median = vapply(
      existing
      , function(col) as.numeric(stats_wide[[paste0(col, "__median")]])
      , numeric(1L)
    )
    , stringsAsFactors = FALSE
  )
  rownames(stats_long) <- NULL
  cat(
    knitr::kable(
      stats_long
      , digits = 2L
      , caption = paste("Measure summary over all rows:", t)
    )
    , sep = "\n"
  )
  cat("\n")

  s <- sample_table(con, t)
  outlier_rows <- lapply(
    existing
    , function(col) {
      vals <- suppressWarnings(as.numeric(s[[col]]))
      info <- flag_iqr(vals)
      data.frame(
        measure = col
        , sample_n = info$n
        , lower_bound = info$lower
        , upper_bound = info$upper
        , flagged = info$flagged
        , stringsAsFactors = FALSE
      )
    }
  )
  cat(
    knitr::kable(
      dplyr::bind_rows(outlier_rows)
      , digits = 2L
      , caption = paste("Interquartile outlier flags in sample:", t)
    )
    , sep = "\n"
  )
  cat("\n")

  for (col in existing) {
    vals <- suppressWarnings(as.numeric(s[[col]]))
    vals <- vals[!is.na(vals)]
    if (length(vals) == 0L) {
      next
    }
    p <- ggplot2::ggplot(
      data.frame(value = vals)
      , ggplot2::aes(x = value)
    ) +
      ggplot2::geom_histogram(bins = 40L, fill = "#9B59B6", color = "#2C3E50") +
      ggplot2::scale_x_continuous(labels = scales::label_number()) +
      ggplot2::labs(title = paste0(t, ": ", col), x = col, y = "Sample count")
    print(apply_plot_theme(p))
  }
}
```

### Date coverage

Span and distinct counts for every date column. The inventory snapshot
should cover the months of 2025 with no gaps. Order milestones should all
sit on or before the current date, since future dated rows are quarantined
at load time instead of entering the facts.

```{r}
#| label: fact-dates
date_cols <- DBI::dbGetQuery(
  con
  , paste0(
    "SELECT "
    , " c.table_name"
    , " , c.column_name"
    , " FROM information_schema.columns AS c"
    , " WHERE c.table_schema = 'core'"
    , " AND c.table_name IN ("
    , paste(paste0("'", FACT_TABLES, "'"), collapse = " , ")
    , ")"
    , " AND c.data_type IN ('date', 'timestamp without time zone'"
    , " , 'timestamp with time zone')"
    , " ORDER BY c.table_name, c.ordinal_position"
  )
)

date_results <- dplyr::bind_rows(lapply(
  seq_len(nrow(date_cols))
  , function(i) {
    t <- date_cols$table_name[[i]]
    col <- date_cols$column_name[[i]]
    sql <- paste0(
      "SELECT "
      , " MIN(", quoted_ident(col), ") AS min_value"
      , " , MAX(", quoted_ident(col), ") AS max_value"
      , " , COUNT(DISTINCT ", quoted_ident(col), ") AS distinct_count"
      , " , COUNT(*) FILTER (WHERE "
      , quoted_ident(col), " IS NULL) AS null_count"
      , " FROM core.", t
    )
    res <- DBI::dbGetQuery(con, sql)
    data.frame(
      table_name = t
      , column_name = col
      , min_value = as.character(res$min_value[[1L]])
      , max_value = as.character(res$max_value[[1L]])
      , distinct_count = as.numeric(res$distinct_count[[1L]])
      , null_count = as.numeric(res$null_count[[1L]])
      , stringsAsFactors = FALSE
    )
  }
))
knitr::kable(date_results, caption = "Date coverage per fact column")

inv_gap <- DBI::dbGetQuery(
  con
  , paste0(
    "SELECT "
    , " COUNT(DISTINCT f.period_month) AS observed_months"
    , " , (EXTRACT(YEAR FROM age(MAX(f.period_month), MIN(f.period_month))) * 12"
    , " + EXTRACT(MONTH FROM age(MAX(f.period_month), MIN(f.period_month)))"
    , " + 1) AS expected_months"
    , " FROM core.fact_inventory AS f"
  )
)
knitr::kable(inv_gap, caption = "Inventory month continuity, observed vs expected")
```

### Correlations

Pairwise plots across numeric measures on one fact, wherever more than one
measure exists. The sample keeps the render fast. Strong links here guide
which drivers later notebooks should model. Weak links warn against naive
assumptions such as spend always tracking clicks.

```{r}
#| label: fact-correlations
#| results: asis
for (t in FACT_TABLES) {
  measures <- intersect(
    MEASURE_COLS[[t]]
    , table_columns(con, t)$column_name
  )
  if (length(measures) < 2L) {
    next
  }
  cat("\n\n#### ", t, "\n\n", sep = "")
  s <- sample_table(con, t, n = 800L)
  num_df <- as.data.frame(lapply(
    measures
    , function(col) suppressWarnings(as.numeric(s[[col]]))
  ))
  names(num_df) <- measures
  num_df <- num_df[, colSums(is.na(num_df)) < nrow(num_df), drop = FALSE]
  if (ncol(num_df) < 2L) {
    cat("\nFewer than two populated measures in sample.\n\n")
    next
  }
  print(GGally::ggpairs(
    num_df
    , title = paste("Measure correlations:", t)
    , progress = FALSE
  ))
}
```

## Quality cross check

Lightweight mirrors of the five quality loops, scoped to `core` and fully
read only. Each reports aggregate counts. A nonzero count is a failed
check. Any mismatch against the loop results needs investigation before
targeted analysis starts, since the loops gate the pipeline and this
notebook must agree with them.

```{r}
#| label: dq-required-text
required_text_cols <- DBI::dbGetQuery(
  con
  , paste0(
    "SELECT "
    , " c.table_name"
    , " , c.column_name"
    , " FROM information_schema.columns AS c"
    , " WHERE c.table_schema = 'core'"
    , " AND c.table_name IN ("
    , paste(
      paste0("'", c(DIM_TABLES, FACT_TABLES), "'")
      , collapse = " , "
    )
    , ")"
    , " AND c.is_nullable = 'NO'"
    , " AND c.data_type IN ('character varying', 'character', 'text')"
    , " ORDER BY c.table_name, c.ordinal_position"
  )
)
required_text_results <- dplyr::bind_rows(lapply(
  seq_len(nrow(required_text_cols))
  , function(i) {
    t <- required_text_cols$table_name[[i]]
    col <- required_text_cols$column_name[[i]]
    sql <- paste0(
      "SELECT COUNT(*) AS failed_rows"
      , " FROM core.", t
      , " WHERE NULLIF(BTRIM(", quoted_ident(col), "), '') IS NULL"
    )
    n <- DBI::dbGetQuery(con, sql)$failed_rows[[1L]]
    data.frame(
      table_name = t
      , column_name = col
      , blank_rows = as.numeric(n)
      , stringsAsFactors = FALSE
    )
  }
))
knitr::kable(
  required_text_results
  , caption = "Required text check, blank rows per column"
)
```

```{r}
#| label: dq-future-dates
future_date_results <- dplyr::bind_rows(lapply(
  seq_len(nrow(date_cols))
  , function(i) {
    t <- date_cols$table_name[[i]]
    col <- date_cols$column_name[[i]]
    sql <- paste0(
      "SELECT COUNT(*) AS failed_rows"
      , " FROM core.", t
      , " WHERE ", quoted_ident(col), " > CURRENT_DATE"
    )
    n <- DBI::dbGetQuery(con, sql)$failed_rows[[1L]]
    data.frame(
      table_name = t
      , column_name = col
      , future_rows = as.numeric(n)
      , stringsAsFactors = FALSE
    )
  }
))
knitr::kable(future_date_results, caption = "Future date check per date column")
```

```{r}
#| label: dq-negative-numerics
NEGATIVE_COLS <- list(
  fact_orders = c("quantity", "unit_price", "unit_cost", "line_total")
  , fact_order_process = c("amount")
  , fact_inventory = c("quantity")
  , fact_campaign_spend = c("impressions", "clicks", "spend")
)
negative_results <- dplyr::bind_rows(unlist(
  lapply(
    names(NEGATIVE_COLS)
    , function(t) {
      cols <- intersect(
        NEGATIVE_COLS[[t]]
        , table_columns(con, t)$column_name
      )
      lapply(
        cols
        , function(col) {
          sql <- paste0(
            "SELECT COUNT(*) AS failed_rows"
            , " FROM core.", t
            , " WHERE ", quoted_ident(col), " < 0"
          )
          n <- DBI::dbGetQuery(con, sql)$failed_rows[[1L]]
          data.frame(
            table_name = t
            , column_name = col
            , negative_rows = as.numeric(n)
            , stringsAsFactors = FALSE
          )
        }
      )
    }
  )
  , recursive = FALSE
))
knitr::kable(negative_results, caption = "Negative numeric check per measure")
```

```{r}
#| label: dq-orphan-fks
orphan_results <- DBI::dbGetQuery(
  con
  , paste0(
    "SELECT "
    , " 'fact_orders leaves customer_key unresolved' AS check_name"
    , " , COUNT(*) AS unresolved_rows"
    , " FROM core.fact_orders AS f"
    , " LEFT JOIN core.dim_customers AS d"
    , " ON d.customer_key = f.customer_key"
    , " WHERE f.customer_key IS NULL OR d.customer_key IS NULL"
    , " UNION ALL "
    , "SELECT "
    , " 'fact_orders leaves product_key unresolved'"
    , " , COUNT(*)"
    , " FROM core.fact_orders AS f"
    , " LEFT JOIN core.dim_products AS d"
    , " ON d.product_key = f.product_key"
    , " WHERE f.product_key IS NULL OR d.product_key IS NULL"
    , " UNION ALL "
    , "SELECT "
    , " 'fact_orders leaves ship_geo_key unresolved'"
    , " , COUNT(*)"
    , " FROM core.fact_orders AS f"
    , " LEFT JOIN core.dim_geo AS g"
    , " ON g.geo_key = f.ship_geo_key"
    , " WHERE f.ship_geo_key IS NULL OR g.geo_key IS NULL"
    , " UNION ALL "
    , "SELECT "
    , " 'fact_orders leaves bill_geo_key unresolved'"
    , " , COUNT(*)"
    , " FROM core.fact_orders AS f"
    , " LEFT JOIN core.dim_geo AS g"
    , " ON g.geo_key = f.bill_geo_key"
    , " WHERE f.bill_geo_key IS NULL OR g.geo_key IS NULL"
    , " UNION ALL "
    , "SELECT "
    , " 'fact_order_process leaves customer_key unresolved'"
    , " , COUNT(*)"
    , " FROM core.fact_order_process AS f"
    , " LEFT JOIN core.dim_customers AS d"
    , " ON d.customer_key = f.customer_key"
    , " WHERE f.customer_key IS NULL OR d.customer_key IS NULL"
    , " UNION ALL "
    , "SELECT "
    , " 'fact_inventory leaves product_key unresolved'"
    , " , COUNT(*)"
    , " FROM core.fact_inventory AS f"
    , " LEFT JOIN core.dim_products AS d"
    , " ON d.product_key = f.product_key"
    , " WHERE f.product_key IS NULL OR d.product_key IS NULL"
    , " UNION ALL "
    , "SELECT "
    , " 'fact_campaign_spend leaves campaign_key unresolved'"
    , " , COUNT(*)"
    , " FROM core.fact_campaign_spend AS f"
    , " LEFT JOIN core.dim_campaign AS d"
    , " ON d.campaign_key = f.campaign_key"
    , " WHERE f.campaign_key IS NULL OR d.campaign_key IS NULL"
  )
)
knitr::kable(orphan_results, caption = "Unresolved foreign keys, NULL counts as unresolved")
```

## Verdict

```{r}
#| label: verdict
failures <- sum(
  sum(dup_results$duplicate_rows, na.rm = TRUE)
  , sum(grain_results$extra_rows, na.rm = TRUE)
  , sum(required_text_results$blank_rows, na.rm = TRUE)
  , sum(future_date_results$future_rows, na.rm = TRUE)
  , sum(negative_results$negative_rows, na.rm = TRUE)
  , sum(orphan_results$unresolved_rows, na.rm = TRUE)
)
cat("Tables profiled:", length(c(DIM_TABLES, FACT_TABLES)), "\n")
cat("Total failed rows across all cross checks:", failures, "\n")
if (failures == 0L) {
  cat(
    paste0(
      "Verdict: data is clean and ready for targeted analysis. "
      , "Notebooks 01 through 07 may proceed from the shapes described above."
    )
    , "\n"
  )
} else {
  cat(
    paste0(
      "Verdict: failures found. Compare against the five quality loops, "
      , "resolve the mismatch, then rerun this notebook before any "
      , "targeted analysis."
    )
    , "\n"
  )
}
```

```{r}
#| label: cleanup
#| include: false
DBI::dbDisconnect(con)
sessionInfo()
```
