---
title: "01 Customer Tenure: SCD2 version patterns"
subtitle: "Which customers churn attributes often and how long versions live"
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
execute:
  warning: false
  message: false
---

## Purpose

This notebook runs the headless tenure script against `dim_customers` and
presents the result. All calculation logic lives in
`r_analysis/scripts/01_customer_tenure.R`. This document only runs it and
charts the outcome, so the script stays the single source of truth. Render
from the repository root with `make notebooks`. The render needs a populated
warehouse and the same `.env` the Python pipeline uses.

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages({
  library(DBI)
  library(dplyr)
  library(ggplot2)
  library(knitr)
})
helper_path <- file.path("..", "r_analysis", "scripts", "db_connect.R")
if (!file.exists(helper_path)) {
  helper_path <- file.path("r_analysis", "scripts", "db_connect.R")
}
source(helper_path)
r_root <- dirname(dirname(helper_path))
output_dir <- file.path(r_root, "output")
figures_dir <- file.path(r_root, "figures")
source(file.path(r_root, "scripts", "01_customer_tenure.R"))
con <- connect_db()
```

```{r}
#| label: run
run_customer_tenure(con, output_dir, figures_dir)
```

## Finding: version counts

Most customers should sit on exactly one version. A long tail of multi
version customers points at unstable source attributes, often address or
region churn. The table below lists the ten most versioned customers by
business key. The chart shows the full distribution.

```{r}
#| label: versions
tenure <- utils::read.csv(
  file.path(output_dir, "customer_tenure_summary.csv")
)
knitr::kable(
  utils::head(tenure, 10L)
  , caption = "Most versioned customers, top ten"
)
knitr::include_graphics(
  file.path(figures_dir, "customer_tenure_hist.png")
)
```

Single version customers cover
`r paste0(round(100 * mean(tenure$version_count == 1L), 1L), "%")`
of the base. When that share drops over time, attribute churn is rising and
downstream as of joins touch more history rows per customer.

## Finding: version lifespan

Short lived versions suggest noisy source feeds rather than real customer
moves. Long lived versions are the stable base the business reports against.

```{r}
#| label: lifespans
lifespan <- utils::read.csv(
  file.path(output_dir, "customer_lifespan_summary.csv")
)
knitr::kable(lifespan, caption = "Version lifespan in days")
knitr::include_graphics(
  file.path(figures_dir, "customer_lifespan_hist.png")
)
```

```{r}
#| label: cleanup
#| include: false
DBI::dbDisconnect(con)
```
