---
title: "02 Revenue Seasonality: monthly pattern and drivers"
subtitle: "Trend and seasonal split of order revenue plus top products"
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
execute:
  warning: false
  message: false
---

## Purpose

This notebook runs the headless seasonality script against `fact_orders`
and presents the result. All calculation logic lives in
`r_analysis/scripts/02_revenue_seasonality.R`. This document only runs it
and charts the outcome, so the script stays the single source of truth.
Render from the repository root with `make notebooks`. The render needs a
populated warehouse and the same `.env` the Python pipeline uses.

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages({
  library(DBI)
  library(dplyr)
  library(ggplot2)
  library(knitr)
})
helper_path <- file.path("..", "r_analysis", "scripts", "db_connect.R")
if (!file.exists(helper_path)) {
  helper_path <- file.path("r_analysis", "scripts", "db_connect.R")
}
source(helper_path)
r_root <- dirname(dirname(helper_path))
output_dir <- file.path(r_root, "output")
figures_dir <- file.path(r_root, "figures")
source(file.path(r_root, "scripts", "02_revenue_seasonality.R"))
con <- connect_db()
```

```{r}
#| label: run
run_revenue_seasonality(con, output_dir, figures_dir)
```

## Finding: monthly revenue with trend and seasonal split

The top panel is observed revenue. The trend panel shows the underlying
direction once repeating effects are removed. The seasonal panel shows the
repeating yearly shape. A flat seasonal panel means no real seasonal pattern
exists and month to month moves are trend plus noise. The split needs at
least two full yearly cycles to settle. Shorter histories report a trend
without a seasonal panel.

```{r}
#| label: monthly
monthly <- utils::read.csv(
  file.path(output_dir, "revenue_seasonality_monthly.csv")
)
knitr::kable(monthly, caption = "Revenue per month")
knitr::include_graphics(
  file.path(figures_dir, "revenue_seasonality_stl.png")
)
```

## Finding: products behind the totals

Seasonality means little without knowing which products drive it. The table
below ranks the top ten products by revenue. When one or two products own a
large share, the seasonal shape is their shape and promo planning should
start with them.

```{r}
#| label: drivers
drivers <- utils::read.csv(
  file.path(output_dir, "revenue_top_products.csv")
)
knitr::kable(drivers, caption = "Top ten products by revenue")
```

```{r}
#| label: cleanup
#| include: false
DBI::dbDisconnect(con)
```
