---
title: "03 Fulfillment Funnel: where orders stall"
subtitle: "Stage conversion from ordered to paid plus time to pay survival"
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
execute:
  warning: false
  message: false
---

## Purpose

This notebook runs the headless funnel script against the accumulating
snapshot `fact_order_process` and presents the result. All calculation logic
lives in `r_analysis/scripts/03_fulfillment_funnel.R`. This document only
runs it and charts the outcome, so the script stays the single source of
truth. Render from the repository root with `make notebooks`. The render
needs a populated warehouse and the same `.env` the Python pipeline uses.

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages({
  library(DBI)
  library(dplyr)
  library(ggplot2)
  library(knitr)
})
helper_path <- file.path("..", "r_analysis", "scripts", "db_connect.R")
if (!file.exists(helper_path)) {
  helper_path <- file.path("r_analysis", "scripts", "db_connect.R")
}
source(helper_path)
r_root <- dirname(dirname(helper_path))
output_dir <- file.path(r_root, "output")
figures_dir <- file.path(r_root, "figures")
source(file.path(r_root, "scripts", "03_fulfillment_funnel.R"))
con <- connect_db()
```

```{r}
#| label: run
run_fulfillment_funnel(con, output_dir, figures_dir)
```

## Finding: funnel from ordered to paid

Each bar counts orders that reached the milestone. Labels show the raw count
plus the share of all ordered rows. The steepest drop between two adjacent
bars is where orders stall. A stall between invoiced and paid points at
collection. A stall between ordered and shipped points at fulfillment
capacity.

```{r}
#| label: funnel
funnel <- utils::read.csv(file.path(output_dir, "funnel_counts.csv"))
knitr::kable(funnel, caption = "Orders per milestone with conversion")
knitr::include_graphics(
  file.path(figures_dir, "fulfillment_funnel.png")
)
```

## Finding: time to pay with censoring

Unpaid orders still count as information: they are censored at the current
date rather than dropped. The curve shows the share of invoiced orders still
unpaid after each day. It falls fast at first and then flattens. The summary
table reports the median days to pay plus the outlier boundary from the
interquartile rule. Orders past that boundary deserve a look as possible
collection problems rather than normal delay.

```{r}
#| label: paydelay
ttp <- utils::read.csv(
  file.path(output_dir, "time_to_pay_summary.csv")
)
knitr::kable(ttp, caption = "Time to pay summary with outlier boundary")
knitr::include_graphics(
  file.path(figures_dir, "pay_survival.png")
)
```

```{r}
#| label: cleanup
#| include: false
DBI::dbDisconnect(con)
```
