---
title: "04 Campaign Lift: spend against revenue"
subtitle: "Whether spend tracks revenue and which campaigns break the pattern"
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
execute:
  warning: false
  message: false
---

## Purpose

This notebook runs the headless lift script and presents the result. Spend
comes from `fact_campaign_spend`. Promoted products come from
`fact_less_fact`. Revenue for those products inside the campaign window
comes from `fact_orders`. The three facts meet only through the shared
dimensions, never directly. All calculation logic lives in
`r_analysis/scripts/04_campaign_lift.R`. This document only runs it and
charts the outcome, so the script stays the single source of truth. Render
from the repository root with `make notebooks`.

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages({
  library(DBI)
  library(dplyr)
  library(ggplot2)
  library(knitr)
})
helper_path <- file.path("..", "r_analysis", "scripts", "db_connect.R")
if (!file.exists(helper_path)) {
  helper_path <- file.path("r_analysis", "scripts", "db_connect.R")
}
source(helper_path)
r_root <- dirname(dirname(helper_path))
output_dir <- file.path(r_root, "output")
figures_dir <- file.path(r_root, "figures")
source(file.path(r_root, "scripts", "04_campaign_lift.R"))
con <- connect_db()
```

```{r}
#| label: run
run_campaign_lift(con, output_dir, figures_dir)
```

## Finding: spend against window revenue

Each point is one campaign. The line is a linear fit with a confidence band.
Point size shows how many SKUs the campaign promoted. Revenue here means
promoted SKU revenue inside the campaign window only, which is a rough
attribution. It credits the campaign with sales that may have happened
anyway, so read the slope as association rather than proof of cause.

```{r}
#| label: lift
lift <- utils::read.csv(file.path(output_dir, "campaign_lift.csv"))
knitr::kable(lift, caption = "Spend, reach, promoted SKUs, and window revenue")
knitr::include_graphics(
  file.path(figures_dir, "campaign_lift.png")
)
```

Labelled points sit more than two standard residuals off the line. High
spend with low revenue marks possible waste. Low spend with high revenue
marks campaigns worth studying and repeating. Campaigns with no promoted
SKU links or no window revenue fall out of the chart entirely, which itself
is a finding about catalogue linkage quality.

```{r}
#| label: cleanup
#| include: false
DBI::dbDisconnect(con)
```
