---
title: "05 Inventory Health: stock by category and ABC split"
subtitle: "Small multiples across months of 2025 plus stockout flags"
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
execute:
  warning: false
  message: false
---

## Purpose

This notebook runs the headless inventory script against the periodic
snapshot `fact_inventory` joined to `dim_products` for category names and
presents the result. All calculation logic lives in
`r_analysis/scripts/05_inventory_health.R`. This document only runs it and
charts the outcome, so the script stays the single source of truth. Render
from the repository root with `make notebooks`. Coverage is the months of
2025 present in the warehouse.

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages({
  library(DBI)
  library(dplyr)
  library(ggplot2)
  library(knitr)
})
helper_path <- file.path("..", "r_analysis", "scripts", "db_connect.R")
if (!file.exists(helper_path)) {
  helper_path <- file.path("r_analysis", "scripts", "db_connect.R")
}
source(helper_path)
r_root <- dirname(dirname(helper_path))
output_dir <- file.path(r_root, "output")
figures_dir <- file.path(r_root, "figures")
source(file.path(r_root, "scripts", "05_inventory_health.R"))
con <- connect_db()
```

```{r}
#| label: run
run_inventory_health(con, output_dir, figures_dir)
```

## Finding: stock per category

One panel per category with its own scale, so a small category trend stays
visible next to a large one. Falling lines into year end flag possible
stockout risk ahead. Rising lines with no matching sales story flag possible
overstock and tied up cash.

```{r}
#| label: multiples
knitr::include_graphics(
  file.path(figures_dir, "inventory_health.png")
)
```

## Finding: flags and ABC split

The flag table compares second half average stock against first half
average stock per category. A drop of thirty percent or more reads as
toward stockout. A rise of fifty percent or more reads as toward overstock.
The ABC split ranks products by total quantity share. Class A products own
the first eighty percent and deserve the tightest replenishment control.
Class C products own the tail and are candidates for review rather than
routine restocking.

```{r}
#| label: flags
flags <- utils::read.csv(file.path(output_dir, "inventory_flags.csv"))
knitr::kable(flags, caption = "Category flags with half year comparison")
abc <- utils::read.csv(file.path(output_dir, "inventory_abc.csv"))
knitr::kable(
  utils::head(abc, 15L)
  , caption = "ABC ranking, top fifteen products"
)
```

```{r}
#| label: cleanup
#| include: false
DBI::dbDisconnect(con)
```
