---
title: "06 Pipeline Health: run duration trends"
subtitle: "Whether the pipeline slows over time and which stage costs most"
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
execute:
  warning: false
  message: false
---

## Purpose

This notebook runs the headless pipeline script against
`core.pipeline_run_log` and presents the result. All calculation logic lives
in `r_analysis/scripts/06_pipeline_health.R`. This document only runs it and
charts the outcome, so the script stays the single source of truth. Render
from the repository root with `make notebooks`. When the log table does not
exist yet, the script skips gracefully and this notebook reports that
instead of failing.

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages({
  library(DBI)
  library(dplyr)
  library(ggplot2)
  library(knitr)
})
helper_path <- file.path("..", "r_analysis", "scripts", "db_connect.R")
if (!file.exists(helper_path)) {
  helper_path <- file.path("r_analysis", "scripts", "db_connect.R")
}
source(helper_path)
r_root <- dirname(dirname(helper_path))
output_dir <- file.path(r_root, "output")
figures_dir <- file.path(r_root, "figures")
source(file.path(r_root, "scripts", "06_pipeline_health.R"))
con <- connect_db()
```

```{r}
#| label: run
run_pipeline_health(con, output_dir, figures_dir)
```

## Finding: duration trend with outliers

The line tracks total run duration over time. The dashed line marks the
outlier boundary from the interquartile rule. Points above it are unusually
slow runs worth comparing against what changed upstream on those dates. A
steady climb of the whole line means the pipeline itself is slowing, which
usually traces back to data growth rather than one bad run.

```{r}
#| label: trend
outlier_path <- file.path(output_dir, "pipeline_outliers.csv")
if (file.exists(outlier_path)) {
  outliers <- utils::read.csv(outlier_path)
  if (nrow(outliers) > 0L) {
    knitr::kable(outliers, caption = "Flagged slow runs")
  } else {
    cat("No runs crossed the outlier boundary.")
  }
  knitr::include_graphics(
    file.path(figures_dir, "pipeline_health.png")
  )
} else {
  cat("No pipeline log found, so no trend exists yet.")
}
```

## Finding: stage cost split

The stacked bar splits the most recent thirty runs into stages. The stage
whose band grows fastest across runs is the growing cost of the pipeline.
That is where optimization effort pays off first.

```{r}
#| label: stages
if (file.exists(outlier_path)) {
  knitr::include_graphics(
    file.path(figures_dir, "pipeline_stages.png")
  )
} else {
  cat("No pipeline log found, so no stage split exists yet.")
}
```

```{r}
#| label: cleanup
#| include: false
DBI::dbDisconnect(con)
```
