---
title: "07 Geo Split: ship to against bill to"
subtitle: "Order volume by region and where the two roles diverge"
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
execute:
  warning: false
  message: false
---

## Purpose

This notebook runs the headless geography script against `fact_orders`
joined twice to the role playing `dim_geo` and presents the result. All
calculation logic lives in `r_analysis/scripts/07_geo_split.R`. This
document only runs it and charts the outcome, so the script stays the single
source of truth. Render from the repository root with `make notebooks`. The
render needs a populated warehouse and the same `.env` the Python pipeline
uses.

```{r}
#| label: setup
#| include: false
suppressPackageStartupMessages({
  library(DBI)
  library(dplyr)
  library(ggplot2)
  library(knitr)
})
helper_path <- file.path("..", "r_analysis", "scripts", "db_connect.R")
if (!file.exists(helper_path)) {
  helper_path <- file.path("r_analysis", "scripts", "db_connect.R")
}
source(helper_path)
r_root <- dirname(dirname(helper_path))
output_dir <- file.path(r_root, "output")
figures_dir <- file.path(r_root, "figures")
source(file.path(r_root, "scripts", "07_geo_split.R"))
con <- connect_db()
```

```{r}
#| label: run
run_geo_split(con, output_dir, figures_dir)
```

## Finding: volume by region per role

Paired bars show ship to volume beside bill to volume for each region.
Regions where the two bars match are locally fulfilled and locally billed.
Regions where they part company buy in one place and receive in another,
which matters for tax, shipping cost, and regional reporting.

```{r}
#| label: split
geo <- utils::read.csv(file.path(output_dir, "geo_ship_bill.csv"))
knitr::kable(geo, caption = "Order lines and revenue per region per role")
knitr::include_graphics(
  file.path(figures_dir, "geo_split.png")
)
```

## Finding: divergence share

The headline number is the share of order lines where ship to region differs
from bill to region. A low share means geography can be treated as one role
in most reporting. A high share means the two roles must stay separate or
regional totals will mislead.

```{r}
#| label: divergence
div <- utils::read.csv(file.path(output_dir, "geo_divergence.csv"))
knitr::kable(div, caption = "Divergent lines and share of total")
```

```{r}
#| label: cleanup
#| include: false
DBI::dbDisconnect(con)
```
